---
title: "Presenting Data with R Markdown"
author: "Jared Sampson"
date: "2019-03-05"
output:
  beamer_presentation: 
    slide_level: 1
    latex_engine: xelatex
monofont: Andale Mono
mainfont: Arial 
subtitle: 'PHAR G8012: Statistics for the Basic Sciences'
header-includes: \usepackage{smartdiagram}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
options(width=40)
```

# A typical scientific workflow
\begin{center}
\smartdiagramset{
    uniform arrow color=true,
    text width=2.25cm,
}
\smartdiagram[circular diagram:clockwise]{
    Collect Data,
    Analyze Data,
    Make Figures,
    Write,
    Design New Experiments
}
\end{center}


# Things we shouldn't have to do when writing a paper

- Scour over a data table to make sure it matches the latest version of the Excel worksheet where the calculations were done, because we can't copy it again or we'd ruin the formatting.

- Completely redo the data analysis if we add a couple more samples.

- Re-check the statistics reported in the paper and any labels added to plots (e.g. significance of results, R^2^ values) after tweaking the analysis procedure.

- Feel like you're spending more time formatting the paper than writing it.



# R Markdown

- Plain-text format, use any text editor

- Embedded R code is evaluated in-place at compile time.

- Go from raw data to final figures in a single step.

- Less copy-pasting/scouring/redoing work you've already done.

- Based on the following open-source projects:

    - [Markdown](https://daringfireball.net/projects/markdown/) (John Gruber)
    - [knitr](https://yihui.name/knitr/) (Yihui Xie, now at RStudio)
    - [pandoc](http://pandoc.org/) (John McFarlane)



# Why use R Markdown?

1. Reproducibility
    - Each analysis step is documented.
    - Source code travels with the text.

2. Automation
    - Save time.
    - Avoid tedious formatting and transcription errors.

3. Durability
    - Plain text is universal.
    - Multiple supported output formats.
    - Community-based support.



# What can you do with R Markdown?

\begin{center}
\smartdiagram[bubble diagram]{
    Text + Data,
    Figures,
    Reports,
    Presentations,
    Papers,
    Grants,
    \emph{Problem Sets!},
    Thesis
}
\end{center}



# General workflow
\begin{center}
\smartdiagram[descriptive diagram]{
    {Setup, {Create a folder (or RStudio project) with your data file.}},
    {Write, {Combine text, R code and figures in an \texttt{.Rmd} file.}},
    {Knit, {\texttt{knitr} compiles R Markdown source to Markdown + figures.
            \texttt{pandoc} converts Markdown + figures to HTML / PDF / Word.}},
    {Repeat, {Easily modify and re-generate your document at any time.}},
}
\end{center}



# A simple R Markdown document

````
## A Level-2 Heading

You can make things **bold** or *italic* (or _italic_).
Or ***both***.
Here's a sentence with a [link](http://www.example.com/).

```{r cars}`r ''`
summary(cars)
```
````



# A simple R Markdown document (knitted)

## A Level-2 Heading

You can make things **bold** or *italic* (or _italic_).
Or ***both***.
Here's a sentence with a [link](http://www.example.com/).

```{r cars}
summary(cars)
```



# YAML "frontmatter" (header)

```
---
title: "Hello World"
author: "Jared Sampson"
date: "2019-03-05"
output: pdf_document
---
```

- [YAML](http://yaml.org/): "a human friendly data serialization standard".

- Goes at the top between 2 lines with `---`.

- Holds document metadata and `pandoc` output settings.



# `knitr` global configuration options 

````
```{r setup, include=FALSE}`r ''`
knitr::opts_chunk$set(echo=TRUE)
```
````

- Set default options to be applied to all code blocks ("chunks") with `opts_chunk$set(...)`.

- Note the `include=FALSE`.  This code does not appear in the document.



# Evaluate R code in code blocks or inline

````
First let's do some calculations in a block:

```{r blocks_or_inline}`r ''`
mean_speed <- mean(cars$speed)
first_100 <- sum(1:100)
```

Then let's explain them in a paragraph.  The mean speed
was `r '\x60r mean_speed\x60'` mph, and the sum of the first 100
positive integers is `r '\x60r first_100\x60'`.

We can also do calculations directly inline.  For 
example, the first 10 numbers of the Fibonacci series 
are `r '\x60r fib <- numeric(10); fib[1] <- 1; fib[2] <- 1;
for (i in 3:10) {fib[i] <- fib[i-2] + fib[i-1]}; fib\x60'`.
````



# Evaluate R code in code blocks or inline (formatted)

First let's do some calculations in a block:

```{r blocks_or_inline}
mean_speed <- mean(cars$speed)
first_100 <- sum(1:100)
```

Then let's explain them in a paragraph.  The mean speed was `r mean_speed` mph, and the sum of the first 100 positive integers is `r first_100`.

We can, of course, also do calculations directly inline.  For example, the first 10 numbers of the Fibonacci series are `r fib <- numeric(10); fib[1] <- 1; fib[2] <- 1; for (i in 3:10) {fib[i] <- fib[i-2] + fib[i-1]}; fib`.


# Include plots

````
Here are 2 histograms from some car data.

```{r hist, echo=FALSE}`r ''`
par(mfrow=c(1, 2))
hist(cars$speed)
hist(cars$dist)
```
````



# Include plots (formatted)

Here are 2 histograms from some car data.

```{r hist, echo=FALSE}
par(mfrow=c(1, 2))
hist(cars$speed)
hist(cars$dist)
```



# Demo

\begin{center}
Let's dive in!
\end{center}