---
title: "R Markdown Demo"
subtitle: "PHAR G8012: Statistics for the Basic Sciences"
author: "Jared Sampson"
date: "2020-02-25"
monofont: Monaco
mainfont: Times New Roman
fontsize: 12pt
output:
  pdf_document: 
    latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
# This R code block is named "setup", and it will not be included in the final document

# Note that I had to employ some tricks below to get R-formatted code blocks not to be
# interpreted as such by the `knitr` parser. This usually won't be necessary for most
# R Markdown documents.

# Basic options for knitr "chunks"
knitr::opts_chunk$set(echo=TRUE, tidy=TRUE)
```

# Before We Begin

The instructions here are for working with R Markdown documents within [RStudio](https://www.rstudio.com/).  If you choose to use another editor or work from the command line (i.e. standalone `R`) you will probably need to make some adjustments to the workflow.

Before creating your first `.Rmd` document, you need to install the R Markdown package into your `R` environment.  This can be done easily by running the command `install.packages("rmarkdown")` at the RStudio console command prompt.  This may take a few moments as dependencies are installed.  Note that RStudio comes bundled with its own `pandoc` executable, which is required for converting the `.md` (Markdown) files produced by `knitr` into `.html`, `.docx`, or `.pdf`.  You can read more about R Markdown in the [documentation](https://rmarkdown.rstudio.com/docs/index.html).

Also note that the `.Rmd` source file for this document is included alongside the PDF, in case you'd like to check it out.



# An R Markdown Exercise

## Create a new RStudio project

1. File > New Project... > Existing Directory... > Browse... (to this `rmarkdown_demo` directory) > Create Project.

1. File > New File > R Markdown...

1. Select 'Document', add Title and select HTML as an output format.


## Inspect your .Rmd document

1. Save the document as **demo.Rmd** to the project folder.

1. Press **Knit** (or `Shift`+`Cmd`+`K` for Mac / `Shift`+`Ctrl`+`K` for Windows) to compile your document.  You should see the formatted HTML document appear in a new RStudio window.

1. Look at the source code to see how it defines e.g. `**bold**` text and "fenced" (```` ``` ````) code blocks.

1. Look at the first (`setup`) code block.  Click the green arrow to run the code in the block.  Note there is no output due to the `include=FALSE` option.

1. Look at the second (`cars`) code block.  Run this code block, too, and see how the output gets printed just below it.  Try the third code block as well.  You should see a plot of pressure vs. temperature.

1. Note that in this instructions document, `R` chunks are shown with a grey background and syntax highlighting, but plain (i.e. non-`R`) code blocks, which don't have the `{r chunk-name}` element directly the opening fence, are shown in a fixed-width font but not shaded grey or syntax-highlighted.  I'll use these plain code blocks to indicate non-`R` text  for your .Rmd document that is not specifically `R` code.



## A "real" problem

You'll find the `bodytempandheartrate.csv` dataset from Courseworks already in this `rmarkdown_demo` project folder.  We're going to do problem #1a from 2018's Problem Set 2: Calculate 95% confidence intervals for body temperature and heart rate for the sample as a whole and for men and women separately.

1. Add a header by typing the following on a line by itself.

    ```
    # Problem 1a.
    ```

    See what happens if you add extra hash/pound/octothorpe symbols and knit again---use 1-6 of these these to get nested heading levels 1 through 6.

1. Add a description of the question you are answering under the header.  Try making it bold.  Where do you have to put the asterisks?

1. Insert an `R` code block via "Insert > R".

1. Write some code in the code block.  Include comments for anything that might be unclear.

    ```{r hr_conf_ints}
    # Import the data
    bthr <- read.csv("bodytempandheartrate.csv")
    
    # Convenience variables
    temp <- bthr$T
    hr <- bthr$HR
    sex <- bthr$MF  # 1 = Male, 2 = Female
    temp_m <- temp[sex == 1]
    temp_f <- temp[sex == 2]
    hr_m <- hr[sex == 1]
    hr_f <- hr[sex == 2]
    
    # Calculate confidence intervals
    temp_all_conf <- t.test(temp)$conf.int
    temp_m_conf <- t.test(temp_m)$conf.int
    temp_f_conf <- t.test(temp_f)$conf.int
    hr_all_conf <- t.test(hr)$conf.int
    hr_m_conf <- t.test(hr_m)$conf.int
    hr_f_conf <- t.test(hr_f)$conf.int
    ```

1. We can access the values calculated for these stored confidence intervals using the form `` `r '\x60r hr_all_conf\x60'` `` inline in a regular line of text.  After the code chunk, add a paragraph or other text (here, a bulleted list, note the blank line between the paragraph and the bullets) to answer the question.

    ```
    The confidence intervals for heart rate are:
    
    * All: `r '\x60r hr_all_conf\x60'`
    * Men: `r '\x60r hr_m_conf\x60'`
    * Women: `r '\x60r hr_f_conf\x60'`
    ```

1. You may have noticed that `R` uses 7 decimal places by default.  That seems like a lot of unnecessary digits.  Let's adjust that output option.  In the `setup` code chunk at the very top of the document, add the following line and re-knit.

    ```{r options, eval=FALSE}
    options(digits=2)
    ```

    Note that `options()` is not specific to R Markdown, but is a general `R` function, so you can use it in any `R` script.


## Make a figure

1. Insert a new R code block and give it a name.

1. Add some histograms or boxplot of the temperature data (always a good idea to do this before even calculating any statistics). 

    ```{r temp_figs, eval=FALSE}
    hist(temp_m, main="Body Temperature (Men)", xlab="°F")
    hist(temp_f, main="Body Temperature (Women)", xlab="°F")
    boxplot(temp_m, temp_f, main="Body Temperature", ylab="°F", names=c("Men", "Women"), notch=TRUE)
    ```
    
1. That takes up too much space.  Let's combine them into one row by adding the following line before the first `hist` statement.

    ```{r par, eval=FALSE}
    par(mfrow=c(1, 3))  # 1x3 matrix figure, populated in row order
    ```

1. But now the aspect ratio is too tall.  Improve it by adding `fig.width` and `fig.asp` options after the block name, so it looks like `{r some_figs, fig.width=3, fig.asp=0.4}`.

1. If we want a floating figure (i.e. not move with the text but be anchored to the top or bottom of a page like a journal figure), add `fig.cap="Some figure caption."` to the block options.  Note that this functions differently depending on the output type (HTML, PDF, Word).


## Other things you may want to try

1. Use the down arrow next to the Knit button to knit a Word document.  Notice the extra output format that appears in the YAML frontmatter (header).  If you want to customize the appearance, edit the styles in the generated .docx document to your liking, save it as e.g. `template.docx`, and add the following to the YAML frontmatter.

    ```
    output:
        word_document:
            reference_docx: path/to/template.docx
    ```
    
    See https://vimeo.com/110804387 for a video example.

1. If you've installed LaTeX, try creating a PDF.  (If not, a lightweight option is [TinyTeX](https://yihui.name/tinytex/), also from Yihui Xie, the creator of `knitr`.)  Perhaps you don't like the default font in the RStudio PDF template, and would like to change the fonts.  You can do so like this within the YAML header:
    
    ```
    output:
        pdf_document:
            latex_engine: xelatex
    mainfont: Times New Roman
    monofont: Courier
    fontsize: 11pt
    ```
    
    This asks Pandoc to generate the PDF using a different LaTeX engine that recognizes the font-related frontmatter options. XeLaTeX is also better able than the default engine to handle non-ASCII characters such as Greek letters.  Although the default engine (`pdflatex`) does handle LaTeX math mode expressions just fine: `$\pi$` turns into $\pi$ ; and the expression
    
        $$ \sum_{n=1}^{\infty} 2^{-n} = 1 $$
    
    results in the following nice-looking equation.
  
    $$ \sum_{n=1}^{\infty} 2^{-n} = 1 $$
  
1. Write a paper or (gasp) your thesis in R Markdown.  It's been done!

- Mike Frank & Chris Hartgerink: [RMarkdown for writing reproducible scientific papers](https://libscie.github.io/rmarkdown-workshop/handout.html)
- Rosanna van Hespen: [Writing your thesis with R Markdown](https://rosannavanhespenresearch.wordpress.com/2016/02/03/writing-your-thesis-with-r-markdown-1-getting-started/)
- Robert Myles McDonnell: [Write your thesis or paper in R Markdown](http://robertmyles.github.io/2016/04/15/write-your-thesis-or-paper-in-r-markdown/)
- Chester Ismay: [thesisdown](https://github.com/ismayc/thesisdown)
- Charles McNamara: [\LaTeX template for Columbia University Dissertations](https://charlesmcnamara.com/latex-template-for-columbia-university-dissertations/)
    
## That's it.  Have fun using R Markdown!